A Sound Deep Embedding of Arbitrary Normal Modal Logics in Coq

### 2022-08-30 paper/paper.tex
Minor fixes (typos, reviewers' suggestions, etc)

%%
%% This is file `sample-sigconf.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `sigconf')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-sigconf.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass command.
\documentclass[sigconf,anonymous]{acmart}

@@

\usepackage{minted}
\usepackage{bussproof}
\usepackage{multirow}

@@

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Ariel Agne da Silveira}
\author{Miguel}
\author{Paulo Torrens}
\author{Karina Roggia}
\author{Rodrigo Ribeiro}

@@

Despite that the original motivation of modal systems was to deal with strict implication, 
for many years modal logic was seen as ``the logic of necessity and possibility''. 
Only after the introduction of Kripke semantics they were seen also as ``simple yet
expressive languages for talking about relational structures''~\cite{Blackburn-ModalLogics}. 
A great multitude of modal systems, with various kinds of modal operators, each of them with 
different meanings or properties, have then been constructed in order to provide effective
formalisms for talking about time, space, knowledge, beliefs, actions, 
obligations, among others.

The interpretation of modal logic simply as ``the logic of necessity'' restricts 
not only its applications but also the range of modal systems. In fact, modal logic 
can also be seen as ``the study of \emph{modalities} -- logical operations that qualify 
assertions about the truth of statements''~\cite{Goldblatt-MathofModality}. In this sense, 
the necessitation rule (from $\varphi$ infers $\Box\varphi$) can be relaxed. This brings 
a number of modal systems known as non-normal, whose semantics was defined only in the 
1970's by Montague~\cite{Montague1970} and Scott~\cite{Scott_AdviceModal1970}.

@@

Although there are some dedicated provers for modal logics~\cite{Gleissner17,Mora11},
we believe that using mature tools, like a proof assistant, is a better option for
verifying facts using modal logics. We justify this choice by: 1) the recent
successful histories on verification of software and
mathematics~\cite{Leroy09,Klein10,Gonthier13,Gonthier08}; 2) There is a growing
community of developers using such tools thanks to 
textbooks on the subject~\cite{Nipkow14,Bertot10,Chlipala13} which allows
newbies to join the game of formal verification and 3) The custom automation
facilities supported by current proof assistants allows that trivial facts can
be proved using decision procedures, relieving the user from writing proof
scripts for formulas that can be checked automatically.

In this context, the present work aims to contribute by constructing a library
that formalize a family of modal logics and its related properties in Coq.
By using such a library, a developer could use all of Coq's resources to prove
modal logics facts.

@@

\begin{itemize}
  \item We define a \emph{deep embedding} of modal logic syntax in Coq and
    define its Kripke semantics.
  \item We model the Hilbert deductive systems as a Coq inductive type that is
    parameterized by different axiom systems like K, D, B, T, K4, K5, S4, S5 and GL
    and prove its soundness with respect to Kripke semantics.
    We also formalize several frame correspondence results.
  \item We show a proof of L\"ob's theorem using the Hilbert deductive system
    developed in our library.
\end{itemize}

@@

Coq is a proof assistant based on the calculus of inductive
constructions (CIC)~\cite{manual_cic}, a higher-order typed
$\lambda$-calculus extended with inductive definitions.  Theorem
proving in Coq follows the ideas of the so-called
``BHK-cor\-res\-pon\-dence''\footnote{Abbreviation of Brouwer, Heyting,
  Kolmogorov, de Bruijn and Martin-L\"of Correspondence. This is also
  known as the Curry-Howard ``isomorphism''.}, where types represent
logical formulas, $\lambda$-terms represent proofs
\cite{Bertot10} and the task of checking if a piece of text is a
proof of a given formula corresponds to checking if the term that
represents the proof has the type corresponding to the given formula.

@@

As a tiny example, consider the task of proving the following 
formula of the propositional logic:
\[
(A \to B)\to (B\to C) \to A \to C
\]
In Coq, such theorem can be expressed as:
\begin{minted}{coq}
Section EXAMPLE.
   Variables A B C : Prop.
   Theorem example : (A -> B) -> (B -> C) -> A -> C.
   Proof.
       intros H H' HA. apply H'. apply H. assumption. 
   Qed.
End EXAMPLE.
\end{minted}
In the previous source code piece, we have defined a Coq section named
\mintinline{coq}{EXAMPLE}\footnote{In Coq, we can use sections to delimit the
  scope of local variables.} which declares variables \mintinline{coq}{A},
\mintinline{coq}{B} and \mintinline{coq}{C} as being propositions (i.e. with type
\mintinline{coq}{Prop}). Tactic \mintinline{coq}{intros} introduces variables
\mintinline{coq}{H}, \mintinline{coq}{H'} and \mintinline{coq}{HA} into the (typing) context,
respectively with types \mintinline{coq}{A -> B}, \mintinline{coq}{B -> C} and
\mintinline{coq}{A} and leaves goal \mintinline{coq}{C} to be proved. Tactic
\mintinline{coq}{apply}, used with a term \mintinline{coq}{t}, generates goal
\mintinline{coq}{P}
when there exists \mintinline{coq}{t: P -> Q} in the typing context and the
current goal is \mintinline{coq}{Q}. Thus, \mintinline{coq}{apply H'} changes the goal
from \mintinline{coq}{C} to \mintinline{coq}{B} and \mintinline{coq}{apply H}
changes the goal to \mintinline{coq}{A}. Tactic \mintinline{coq}{assumption}
traverses the typing context to find a hypothesis that matches the goal.

We define next a proof of the previous propositional logical formula
that, in contrast to the previous proof, that was built using tactics
(\mintinline{coq}{intros}, \mintinline{coq}{apply} and \mintinline{coq}{assumption}), is coded
directly as a function:
\begin{minted}{coq}
Definition example :(A -> B) -> (B -> C) -> A -> C :=
 fun (H : A -> B)
     (H' : B -> C)
     (HA : A) => H' (H HA).
\end{minted}
However, even for very simple theorems, coding a definition directly
as a Coq term can be a hard task. Because of this, the use of tactics
has become the standard way of proving theorems in Coq. Furthermore,
the Coq proof assistant provides not only a great number of tactics
but also a domain-specific language for scripted proof automation,
called $\mathcal{L}$tac. Details about $\mathcal{L}$tac and Coq can be found
in~\cite{Chlipala13,Bertot10,manual_coq}.

@@

\paragraph{Syntax and semantics of modal logics}
To implement our library, we use a \emph{deep embedding} of
formulas, i.e. we represent formulas as the following Coq inductive datatype.
\begin{minted}{coq}
Inductive form : Set :=
| Lit     : nat          -> form
| Neg     : form -> form
| Box     : form -> form
| Dia     : form -> form
| And     : form -> form -> form
| Or      : form -> form -> form
| Implies : form -> form -> form.
\end{minted}
Most constructors of the type have an immediate meaning, since they represent
connectives of propositional logics. The constructor \mintinline{coq}{Lit}
is used to construct propositional variables; constructor \mintinline{coq}{Box}
denotes the modal operator $\Box$ and \mintinline{coq}{Dia}, the operator
$\Diamond$.

Before defining the Kripke semantics, we need to represent the model used to
interpret a modal logic formula. We encode a frame by a record formed by a
a type \mintinline{coq}{W : Set} to represent the set of worlds and the
accessibility relation is denoted by a type \mintinline{coq}{R : W -> W -> Prop}.
\begin{minted}{coq}
Record Frame : Type := {
  W : Set;
  R : W -> W -> Prop;
}.
\end{minted}
A model is denoted by another record formed by a frame and a labeling function,
which assign the truth value of a variable at a given world.
\begin{minted}{coq}
Record Model : Type := {
  F : Frame; 
  v : nat -> (W F) -> Prop; 
}.
\end{minted}
Note that the type of the labeling function \mintinline{coq}{v} ensures that it
refers to a world from the set of possible worlds in the frame
\mintinline{coq}{F}. The encoding of the Kripke semantics is done by a Coq function
that implements the relation of Figure~\ref{fig:kripke}. We omit the code
for propositional logics connectives for brevity. The interesting parts are for
variables (constructor \mintinline{coq}{Lit}) which uses the labeling function
\mintinline{coq}{v} from the model \mintinline{coq}{M}. The meaning of
modalities $\Box$ and $\Diamond$ are represented by a quantifying over worlds
using the accessibility relation $R$. 
\begin{minted}{coq}
Fixpoint formula_valuation (M: Model)
                           (w: W (F M))
                           (phi: form): Prop :=
  match phi with
  | Lit     x            => v M x w 
  | Box     psi
     => forall w': W (F M), R (F M) w w' ->
                            formula_valuation M w' psi
  | Dia     psi
     => exists w': W (F M), R (F M) w w' /\
                   formula_valuation M w' psi
  (** some code omitted *)
  end.
\end{minted}
In order to ease the task of writing formulas and its semantics properties, we
define Coq notations that mimics the standard way of writing them.

\paragraph{The Hilbert deductive system}
Basically, a Hilbert-style deductive system allows the use of four
possibilities in a proof: 1) the use of a hypothesis; 2) the use of an axiom;
3) the use of the modus ponens rule or 4) the use of the generalization (necessity)
rule. The inductive type \mintinline{coq}{deduction} encodes these possibilities
as its constructors. The \mintinline{coq}{deduction} type is parameterized by
axioms that form a system for modal logics. In this way, our definition can be
used by several systems of modal logic.
\begin{minted}{coq}
Inductive deduction (A: axiom -> Prop)
                           : theory -> form -> Prop :=
  (* Premise. *)
  | Prem: forall (t: theory)
                 (f: form)
                 (i: nat),
          (nth_error t i = Some f) -> deduction A t f
  (* Axiom. *)
  | Ax: forall (t: theory)
               (a: axiom)
               (f: form),
        A a -> instantiate a = f -> deduction A t f
  (* Modus Ponens. *)
  | Mp: forall (t: theory)
               (f g: form)
               (d1: deduction A t (f -> g))
               (d2: deduction A t f),
        deduction A t g
  (* Generalization. *)
  | Nec: forall (t: theory)
                (f: form)
                (d1: deduction A t f),
         deduction A t (Box f).
\end{minted}
The constructors \mintinline{coq}{Prem}, \mintinline{coq}{Mp} and \mintinline{coq}{Nec}
represent rules for hypothesis, modus ponens and generalization, respectively.
Before discussing the \mintinline{coq}{Ax}, we need to discuss how axioms are
represented in our library. The inductive type \mintinline{coq}{axiom} specifies
the different axioms supported by our library.
\begin{minted}{coq}
Inductive axiom : Set :=
  | ax1   : form -> form -> axiom
  | ax2   : form -> form -> form  -> axiom
  (** some code omitted *)
  | axK   : form -> form -> axiom
  | axPos : form -> form -> axiom
  | axT   : form -> axiom
  | axB   : form -> axiom
  | axK4  : form -> axiom
  | axD   : form -> axiom
  | axK5  : form -> axiom.
\end{minted}
Each constructor of type \mintinline{coq}{axiom} represents an axiom and
its type stores the subformulas which compose the axiom. The type has
17 constructors: the first 10 consists encode axioms for propositional logics
and the last 7 are for different formulas taken as axioms by systems of modal logics.
Axioms are built using the function \mintinline{coq}{instantiate}, which
receives an axiom and builds its corresponding formula by pattern matching as
follows:
\begin{minted}{coq}
Definition instantiate (a: axiom): modalFormula :=
  match a with
  | ax1    p   q
     => [! p -> (q -> p) !]
  | ax2    p   q   r
     => [! (p -> (q -> r)) -> ((p -> q) -> (p -> r)) !]
  | ax3    p   q
     => [! (~ q -> ~ p) -> (p -> q) !]
  | axK    p   q
     => [! [] (p -> q) -> ([] p -> [] q) !]
  | axPos  p   q
     => [! <> (p \/ q) -> (<> p \/ <> q) !]
  (** some code omitted *)
  end.
\end{minted}
Notation \mintinline{coq}{[! p !]} is used to write modal logic formulas using a
syntax closer to mathematical practice using Coq. This is necessary to avoid
parsing ambiguities, since some symbols are also used by Coq's syntax.
Modalities $\Box$ and $\Diamond$ are denoted by \mintinline{coq}{[]}
and \mintinline{coq}{<>}, respectively.

The \mintinline{coq}{Ax} constructor uses the \mintinline{coq}{instantiate}
function to create a formula instance from a given system,
represented as a predicate
\mintinline{coq}{A : axiom -> Prop} which parameterizes our deduction type. An
axiom \mintinline{coq}{a} is valid within a system \mintinline{coq}{A} whenever
the type \mintinline{coq}{A a} is inhabited, i.e. it is not empty. Next, we
describe how to represent different systems of modal logic.

\paragraph{Systems of modal logics} In order to use the deductive system to
check the provability of formulas, we need to specify the axioms that are
included in the system. Our approach is to define predicates
which denote each system. The simplest set of axioms is usually refereed as
\textbf{K} and it includes the axiom $\Box(A \to B) \to (\Box A \to \Box B)$
and all the axioms for propositional logic. We represent system \textbf{K}
by the following inductive type:
\begin{minted}{coq}
Inductive K: axiom -> Prop :=
  | K_ax1: forall p q, K (ax1 p q)
  | K_ax2: forall p q r, K (ax2 p q r)
  (** some code omitted *)
  | K_axK: forall p q, K (axK p q)
\end{minted}
Type \mintinline{coq}{K} is a predicate over axioms: it has one constructor for
each of the valid axioms which form the system. Given a value
of \mintinline{coq}{axiom}, a value of type \mintinline{coq}{K a} denotes a
proof that a given axiom is part of the system \textbf{K}. This is the main
trick used by the constructor \mintinline{coq}{Ax}, of type
\mintinline{coq}{deduction}, to generically represent axioms for
different systems of modal logics uniformly. In order to understand how it
works, let us review its type (below, remember that \mintinline{coq}{A : axiom -> Prop}
denotes a system of axioms for modal logics):
\begin{minted}{coq}
Ax: forall (t: theory)
           (a: axiom)
           (f: form),
           A a -> instantiate a = f -> deduction A t f
\end{minted}
Note that the first parameter of constructor \mintinline{coq}{Ax} is a value
of type \mintinline{coq}{A a}, which is only inhabited if the predicate
\mintinline{coq}{A} holds for the axiom \mintinline{coq}{a}. Through this
evidence, is possible to create the corresponding axiom formula instance by
using function \mintinline{coq}{instantiate}. In this way, we are able to
represent all systems of axioms for modal logics using a single constructor
for its axioms.

@@

A Coq formalization for modal systems $K$, $T$ and $K4$ was the subject of 
Doczkal and Smolka's work. Author's interest was in the definition of dynamic
and temporal logics. The development of the code establishes the construction
and definitions of tactics, through lemmas propositional operators and relation
properties were constructed. Each model was defined through structures, which
consist of worlds, relationships and propositions~\cite{saarland}.

@@

The next table summarizes the formalized modal logics systems by our library
and related works.
\vspace{0.5cm} {
  \begin{table}[H]
    \begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}
      \hline
      \multirow{2}{*}{Authors} & \multicolumn{9}{c|}{Systems considered} \\ \cline{2-10} 
                               & K & D & B & T & K4 & K5 & S4 & S5 & GL\\ \hline
      Wind et. al.~\cite{dewind}     & X &   &   & X & X  & X  & X  &  X & \\ \hline
      Doczkal et. al.~\cite{saarland}   & X &   &   & X & X  &    &    &  &  \\ \hline
      Benzmüller et. al.~\cite{Chris_Bruno}& X &   & X & X &  X &    &    & X &  \\ \hline
      Bentzen~\cite{Bentzen2021} & X & & & X & & X & X & X & \\ \hline
      Maggesi et. al.~\cite{Maggesi21} & X & & & & & & & & X \\ \hline
      This work         & X & X & X & X & X  &  X &  X & X  & X \\ \hline
    \end{tabular}
    \centering
    \caption{Comparision with different Coq modal logics developements.}
    \label{tab:comparacao_trabalhos_rel}
  \end{table}
}

As we can see, as far we know, our work is the unique formalization that covers systems
K, D, B, T, K4, K5, S4, S5 and GL. Also, since our deduction is
parameterized by system of axioms considered, it can be used to represent any
system of normal modal logics, unlike any other formalizations found in the literature.

\paragraph{Applications of modal logics in semantics} Another application of
modal logics is its use for reasoning about programming languages semantics.
Dynamic logic was introduced by Pratt in late 70's as modal logic for reasoning
about programs~\cite{Pratt76,Pratt80}. The key idea on dynamic logic is to express the
Hoare triples by a modality meaning ``after a program $p$ executes, formula
$\varphi$ will be true''. The programming languages considered in Pratt's
original work are composed by an assignment and regular language operators
(union, concatenation and Kleene star)~\cite{Pratt76}.

>>>>

%%
%% This is file `sample-sigconf.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `sigconf')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-sigconf.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass command.
\documentclass[sigconf]{acmart}

@@

\usepackage{minted}
\usepackage{bussproofs}
\usepackage{multirow}

@@

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.

\author{Ariel Agne da Silveira}
\affiliation{\institution{Federal University of Ouro Preto}\country{}}
\email{ariel.silveira@aluno.ufop.edu.br}

\author{Miguel Nunes}
\affiliation{\institution{Santa Catarina State University}\country{}}
\email{miguel.alfredo.nunes@gmail.com}

\author{Karina Roggia}
\affiliation{\institution{Santa Catarina State University}\country{}}
\email{karina.roggia@udesc.br}

\author{Paulo Torrens}
\affiliation{\institution{Santa Catarina State University}\country{}}
\email{paulotorrens@gnu.org}

\author{Rodrigo Ribeiro}
\affiliation{\institution{Federal University of Ouro Preto}\country{}}
\email{rodrigo.ribeiro@ufop.edu.br}

@@

Despite that the original motivation of modal systems was to deal with strict implication, 
for many years modal logic was seen as ``the logic of necessity and possibility''. 
Only after the introduction of Kripke semantics they were also seen as ``simple yet
expressive languages for talking about relational structures''~\cite{Blackburn-ModalLogics}. 
A great multitude of modal systems, with various kinds of modal operators, each of them with 
different meanings or properties, have then been constructed in order to provide effective
formalisms for talking about time, space, knowledge, beliefs, actions, 
obligations, among others.

The interpretation of modal logic simply as ``the logic of necessity'' restricts 
not only its applications but also the range of modal systems. In fact, modal logic 
can also be seen as ``the study of \emph{modalities} -- logical operations that qualify 
assertions about the truth of statements''~\cite{Goldblatt-MathofModality}. In this sense, 
the necessitation rule (from $\varphi$ infer $\Box\varphi$) may be relaxed. This brings 
a number of modal systems known as non-normal, whose semantics were defined only in the 
1970's by Montague~\cite{Montague1970} and Scott~\cite{Scott_AdviceModal1970}.

@@

Although there are some dedicated provers for modal logics~\cite{Gleissner17,Mora11},
we believe that using mature tools, like a proof assistant, is a better option for
verifying facts using modal logics. We justify this choice by: 1) the recent
successful histories on verification of software and
mathematics~\cite{Leroy09,Klein10,Gonthier13,Gonthier08}; 2) there is a growing
community of developers using such tools thanks to 
textbooks on the subject~\cite{Nipkow14,Bertot10,Chlipala13} which allows
newcomers to join the game of formal verification; and 3) the custom automation
facilities supported by current proof assistants allows for trivial facts to
be proven using decision procedures, relieving the user from writing proof
scripts for formulas that can be checked automatically.

In this context, the present work aims to contribute by constructing a library
that formalizes a family of normal modal logics and their related properties in Coq.
By using such a library, a developer may use all of Coq's resources to prove
modal logics facts.

@@

\begin{itemize}
  \item We define a \emph{deep embedding} of modal logic syntax in Coq and
    define its Kripke semantics.
  \item We model the Hilbert deductive systems as a Coq inductive type that is
    parameterized by different axiom systems like K, D, B, T, K4, K5, S4, S5 and GL,
    and prove its soundness with respect to Kripke semantics.
    We also formalize several frame correspondence results.
  \item We show a proof of L\"ob's theorem using the Hilbert deductive system
    developed in our library.
\end{itemize}

@@

Coq is a proof assistant based on the calculus of inductive
constructions (CIC)~\cite{manual_cic}, a higher-order typed
$\lambda$-calculus extended with inductive definitions.  Theorem
proving in Coq follows the ideas of the so-called
``BHK-cor\-res\-pon\-dence''\footnote{Abbreviation of Brouwer, Heyting,
  Kolmogorov, de Bruijn and Martin-L\"of Correspondence. This is also
  known as the Curry-Howard isomorphism.}, where types represent
logical formulas, $\lambda$-terms represent proofs
\cite{Bertot10} and the task of checking if a piece of text is a
proof of a given formula corresponds to checking if the term that
represents the proof has the type corresponding to the given formula.

@@

As a tiny example, consider the task of proving the following 
formula of the propositional logic:
\[
(A \to B)\to (B\to C) \to A \to C
\]
In Coq, such theorem can be expressed as:
\begin{minted}{coq}
Section EXAMPLE.
   Variables A B C: Prop.
   Theorem example: (A -> B) -> (B -> C) -> A -> C.
   Proof.
       intros H H' HA. apply H'. apply H. assumption. 
   Qed.
End EXAMPLE.
\end{minted}
In the previous source code piece, we have defined a Coq section named
\mintinline{coq}{EXAMPLE}\footnote{In Coq, we can use sections to delimit the
  scope of local variables.} which declares variables \mintinline{coq}{A},
\mintinline{coq}{B} and \mintinline{coq}{C} as being propositions (i.e. with type
\mintinline{coq}{Prop}). Tactic \mintinline{coq}{intros} introduces variables
\mintinline{coq}{H}, \mintinline{coq}{H'} and \mintinline{coq}{HA} into the (typing) context,
respectively with types \mintinline{coq}{A -> B}, \mintinline{coq}{B -> C} and
\mintinline{coq}{A} and leaves goal \mintinline{coq}{C} to be proved. Tactic
\mintinline{coq}{apply}, used with a term \mintinline{coq}{t}, generates goal
\mintinline{coq}{P}
when there exists \mintinline{coq}{t: P -> Q} in the typing context and the
current goal is \mintinline{coq}{Q}. Thus, \mintinline{coq}{apply H'} changes the goal
from \mintinline{coq}{C} to \mintinline{coq}{B} and \mintinline{coq}{apply H}
changes the goal to \mintinline{coq}{A}. Tactic \mintinline{coq}{assumption}
traverses the typing context to find a hypothesis that matches the goal.

We define next a proof of the previous propositional logical formula
that, in contrast to the previous proof, that was built using tactics
(\mintinline{coq}{intros}, \mintinline{coq}{apply} and \mintinline{coq}{assumption}), is coded
directly as a function:
\begin{minted}{coq}
Definition example :(A -> B) -> (B -> C) -> A -> C :=
 fun (H: A -> B)
     (H': B -> C)
     (HA: A) => H' (H HA).
\end{minted}
However, even for very simple theorems, coding a definition directly
as a Coq term can be a hard task. Because of this, the use of tactics
has become the standard way of proving theorems in Coq. Furthermore,
the Coq proof assistant provides not only a great number of tactics
but also a domain-specific language for scripted proof automation,
called $\mathcal{L}$tac. Details about $\mathcal{L}$tac and Coq can be found
in~\cite{Chlipala13,Bertot10,manual_coq}.

@@

\paragraph{Syntax and semantics of modal logics}
To implement our library, we use a \emph{deep embedding} of
formulas, i.e. we represent formulas as the following Coq inductive datatype.
\begin{minted}{coq}
Inductive form: Set :=
| Lit    : nat          -> form
| Neg    : form -> form
| Box    : form -> form
| Dia    : form -> form
| And    : form -> form -> form
| Or     : form -> form -> form
| Implies: form -> form -> form.
\end{minted}
Most constructors of the type have an immediate meaning, since they represent
connectives of propositional logics. The constructor \mintinline{coq}{Lit}
is used to construct propositional variables; constructor \mintinline{coq}{Box}
denotes the modal operator $\Box$ and \mintinline{coq}{Dia}, the operator
$\Diamond$.

Before defining the Kripke semantics, we need to represent the model used to
interpret a modal logic formula. We encode a frame by a record formed by
a type \mintinline{coq}{W: Set} to represent the set of worlds and an
accessibility relation denoted by a type \mintinline{coq}{R: W -> W -> Prop}.
\begin{minted}{coq}
Record Frame: Type := {
  W: Set;
  R: W -> W -> Prop;
}.
\end{minted}
A model is then denoted by another record, formed by a frame and a labeling function,
which assign the truth value of a variable at a given world.
\begin{minted}{coq}
Record Model: Type := {
  F: Frame; 
  v: nat -> (W F) -> Prop; 
}.
\end{minted}
Note that the type of the labeling function \mintinline{coq}{v} ensures that it
refers to a world from the set of possible worlds in the frame
\mintinline{coq}{F} (we recall that \mintinline{coq}{W F} is Coq's syntax for
accessing field \mintinline{coq}{W} from the record \mintinline{coq}{F}). The encoding of the Kripke semantics is done by a Coq function
that implements the relation of Figure~\ref{fig:kripke}. We omit the code
for propositional logics connectives for brevity. The interesting parts are for
variables (constructor \mintinline{coq}{Lit}) which uses the labeling function
\mintinline{coq}{v} from the model \mintinline{coq}{M}. The meaning of
modalities $\Box$ and $\Diamond$ are represented by a quantifying over worlds
using the accessibility relation $R$. 
\begin{minted}{coq}
Fixpoint formula_valuation (M: Model) (w: W (F M))
                                 (phi: form): Prop :=
  match phi with
  | Lit     x
     => v M x w (* should be inhabited, i.e., true *)
  | Box     psi
     => forall w': W (F M), R (F M) w w' ->
                            formula_valuation M w' psi
  | Dia     psi
     => exists w': W (F M), R (F M) w w' /\
                            formula_valuation M w' psi
  (* some code omitted *)
  end.
\end{minted}
In order to ease the task of writing formulas and its semantics properties, we
define Coq notations that mimics the standard way of writing them.

\paragraph{The Hilbert deductive system}
Basically, a Hilbert-style deductive system allows the use of four
possibilities in a proof: 1) the use of a hypothesis; 2) the use of an axiom;
3) the use of the modus ponens rule or 4) the use of the generalization (necessity)
rule. The inductive type \mintinline{coq}{deduction} encodes these possibilities
as its constructors. The \mintinline{coq}{deduction} type is parameterized by
axioms that form a system for modal logics. In this way, our definition can be
used by several systems of modal logic.
\begin{minted}{coq}
Inductive deduction (A: axiom -> Prop)
                           : theory -> form -> Prop :=
  (* Premise. *)
  | Prem: forall (t: theory)
                 (f: form)
                 (i: nat),
          (nth_error t i = Some f) -> deduction A t f
  (* Axiom. *)
  | Ax: forall (t: theory)
               (a: axiom)
               (f: form),
        A a -> instantiate a = f -> deduction A t f
  (* Modus Ponens. *)
  | Mp: forall (t: theory)
               (f g: form)
               (d1: deduction A t (f -> g))
               (d2: deduction A t f),
        deduction A t g
  (* Generalization. *)
  | Nec: forall (t: theory)
                (f: form)
                (d1: deduction A t f),
         deduction A t (Box f).
\end{minted}
The constructors \mintinline{coq}{Prem}, \mintinline{coq}{Mp} and \mintinline{coq}{Nec}
represent rules for hypothesis, modus ponens and generalization, respectively.
Before discussing the \mintinline{coq}{Ax}, we need to discuss how axioms are
represented in our library. The inductive type \mintinline{coq}{axiom} specifies
the different axioms supported by our library.
\begin{minted}{coq}
Inductive axiom: Set :=
  | ax1  : form -> form -> axiom
  | ax2  : form -> form -> form  -> axiom
  (* some code omitted *)
  | axK  : form -> form -> axiom
  | axPos: form -> form -> axiom
  | axT  : form -> axiom
  | axB  : form -> axiom
  | axK4 : form -> axiom
  | axD  : form -> axiom
  | axK5 : form -> axiom.
\end{minted}
Each constructor of type \mintinline{coq}{axiom} represents an axiom and
its type stores the subformulas which compose the axiom. The type has
17 constructors: the first 10 consists encode axioms for propositional logics
and the last 7 are for different formulas taken as axioms by systems of modal logics.
Axioms are built using the function \mintinline{coq}{instantiate}, which
receives an axiom and builds its corresponding formula by pattern matching as
follows:
\begin{minted}{coq}
Definition instantiate (a: axiom): modalFormula :=
  match a with
  | ax1    p   q
     => [! p -> (q -> p) !]
  | ax2    p   q   r
     => [! (p -> (q -> r)) -> ((p -> q) -> (p -> r)) !]
  | ax3    p   q
     => [! (~ q -> ~ p) -> (p -> q) !]
  | axK    p   q
     => [! [] (p -> q) -> ([] p -> [] q) !]
  | axPos  p   q
     => [! <> (p \/ q) -> (<> p \/ <> q) !]
  (* some code omitted *)
  end.
\end{minted}
Notation \mintinline{coq}{[! p !]} is used to write modal logic formulas using a
syntax closer to mathematical practice using Coq. This is necessary to avoid
parsing ambiguities, since some symbols are also used by Coq's syntax.
Modalities $\Box$ and $\Diamond$ are denoted by \mintinline{coq}{[]}
and \mintinline{coq}{<>}, respectively.

The \mintinline{coq}{Ax} constructor uses the \mintinline{coq}{instantiate}
function to create a formula instance from a given system,
represented as a predicate
\mintinline{coq}{A: axiom -> Prop} which parameterizes our deduction type. An
axiom \mintinline{coq}{a} is valid within a system \mintinline{coq}{A} whenever
the type \mintinline{coq}{A a} is inhabited, i.e. it is not empty. Next, we
describe how to represent different systems of modal logic.

\paragraph{Systems of modal logics} In order to use the deductive system to
check the provability of formulas, we need to specify the axioms that are
included in the system. Our approach is to define predicates
which denote each system. The simplest set of axioms is usually refereed as
\textbf{K} and it includes the axiom $\Box(A \to B) \to (\Box A \to \Box B)$
and all the axioms for propositional logic. We represent system \textbf{K}
by the following inductive type:
\begin{minted}{coq}
Inductive K: axiom -> Prop :=
  | K_ax1: forall p q, K (ax1 p q)
  | K_ax2: forall p q r, K (ax2 p q r)
  (* some code omitted *)
  | K_axK: forall p q, K (axK p q)
\end{minted}
Type \mintinline{coq}{K} is a predicate over axioms: it has one constructor for
each of the valid axioms which form the system. Given a value
of \mintinline{coq}{axiom}, a value of type \mintinline{coq}{K a} denotes a
proof that a given axiom is part of the system \textbf{K}. This is the main
trick used by the constructor \mintinline{coq}{Ax}, of type
\mintinline{coq}{deduction}, to generically represent axioms for
different systems of modal logics uniformly. In order to understand how it
works, let us review its type (below, remember that \mintinline{coq}{A: axiom -> Prop}
denotes a system of axioms for modal logics):
\begin{minted}{coq}
Ax: forall (t: theory)
           (a: axiom)
           (f: form),
           A a -> instantiate a = f -> deduction A t f
\end{minted}
Note that the first parameter of constructor \mintinline{coq}{Ax} is a value
of type \mintinline{coq}{A a}, which is only inhabited if the predicate
\mintinline{coq}{A} holds for the axiom \mintinline{coq}{a}. Through this
evidence, is possible to create the corresponding axiom formula instance by
using function \mintinline{coq}{instantiate}. In this way, we are able to
represent all systems of axioms for modal logics using a single constructor
for its axioms.

@@

A Coq formalization for modal systems $K$, $T$ and $K4$ was the subject of 
Doczkal and Smolka's work. The authors' interest was in the definition of dynamic
and temporal logics. The development of the code establishes the construction
and definitions of tactics, through lemmas propositional operators and relation
properties were constructed. Each model was defined through structures, which
consist of worlds, relationships and propositions~\cite{saarland}.

@@

The following table summarizes the formalized modal logics systems by our library
and related works:
\vspace{0.5cm} {
  \begin{table}[H]
    \begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}
      \hline
      \multirow{2}{*}{Authors} & \multicolumn{9}{c|}{Systems considered} \\ \cline{2-10} 
                               & K & D & B & T & K4 & K5 & S4 & S5 & GL\\ \hline
      Wind et. al.~\cite{dewind}     & X &   &   & X & X  & X  & X  &  X & \\ \hline
      Doczkal et. al.~\cite{saarland}   & X &   &   & X & X  &    &    &  &  \\ \hline
      Benzmüller et. al.~\cite{Chris_Bruno}& X &   & X & X &  X &    &    & X &  \\ \hline
      Bentzen~\cite{Bentzen2021} & X & & & X & & X & X & X & \\ \hline
      Maggesi et. al.~\cite{Maggesi21} & X & & & & & & & & X \\ \hline
      This work         & X & X & X & X & X  &  X &  X & X  & X \\ \hline
    \end{tabular}
    \centering
    \caption{Comparision with different Coq modal logics developements.}
    \label{tab:comparacao_trabalhos_rel}
  \end{table}
}

As we can see, as far we know, our work is the only formalization that covers systems
K, D, B, T, K4, K5, S4, S5 and GL. Also, since our deduction is
parameterized by system of axioms considered, it can be used to represent any
system of normal modal logics, unlike any other formalizations found in the literature.

\paragraph{Applications of modal logics in semantics} Another application of
modal logics is its use for reasoning about programming languages semantics.
Dynamic logic was introduced by Pratt in late 70's as modal logic for reasoning
about programs~\cite{Pratt76,Pratt80}. The key idea on dynamic logic is to express the
Hoare triples by a modality meaning ``after a program $p$ executes, formula
$\varphi$ will be true''. The programming languages considered in Pratt's
original work are composed by assignment and regular language operators
(union, concatenation and Kleene star)~\cite{Pratt76}.
